---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# pepdiff

<!-- badges: start -->
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
[![R-CMD-check](https://github.com/TeamMacLean/pepdiff/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/TeamMacLean/pepdiff/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

Workflows for Characterising and Analysing Peptide Quantifications

Provides functions for developing workflows to characterise, evaluate and perform differential abundance analysis in tabular data from quantified Mass Spec experiments. Typically used for experiments with Gene, Peptide, Time, Genotype, Biorep, Techrep and Quantification variables.

## Installation

Install direct from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("TeamMacLean/pepdiff")
```

Or via `renv`.

``` r
# install.packages("devtools")
renv::install("TeamMacLean/pepdiff)
```

## Use

The package contains vignettes describing the basic workflows and you should read them before using the 
package.

There are four vignettes and they should be read in order as they follow the analysis from start to finish.

### Data Checks

How to check the coverage in the data and assess things like it's distribution.

```{r, eval=FALSE}
library(pepdiff)
vignette('data-checks', package="pepdiff")
```


### Doing Comparisons

How to perform comparisons with selected tests


```{r, eval=FALSE}
library(pepdiff)
vignette('doing-comparisons', package="pepdiff")
```


### Check Results

How to evaluate the tests you've chosen and make wiser interpretations and experimental designs

```{r, eval=FALSE}
library(pepdiff)
vignette('check-results', package="pepdiff")
```

### Analysis

How to perform an analysis of a broad set of comparisons

```{r, eval=FALSE}
library(pepdiff)
vignette('analysis', package="pepdiff")
```
