---
title: "basic_use"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{basic_use}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(here)
library(pepdiff)
```

## Load in data
```{r}
d <- import_data(here("inst", "extdata", "anon.csv"),
                 gene_id = "gene_name",
                 peptide = "peptide_sequence",
                 treatment = "treatment_name"
                 )

```

## Check number of missing data points

Includes at tech rep level

Missing data

```{r}
missing_peptides_plot(d)
```

Replicate level
```{r}

times_measured(d)
times_measured_plot(d)
```


## Look at similarity of samples


```{r}
plot_pca(d)

plot_kmeans(d)
```

## Look at distribution of quantifications

```{r}
plot_quant_distributions(d)
plot_quant_distributions(d, log = TRUE)
norm_qqplot(d)
norm_qqplot(d, log = TRUE)
```


## Do a comparison

665e6428 0 seconds vs 665e6428 150 seconds

```{r}
r <- compare(d, iters = 1000, metrics=c("bootstrap_t", "wilcoxon", "rank_product"), control = '665e6428', c_seconds = '0', treatment = '665e6428', t_seconds = '150')

r
```



### Look at distribution of calculated Fold Changes

```{r, eval = TRUE}
plot_fc(r)
fc_qqplot(r)
plot_fc(r, log = TRUE)
fc_qqplot(r, log = TRUE)
```



## Plot p-values

```{r, eval=TRUE, fig.width=12, fig.height=12}
plot_result(r)
```


## Test distribution of p-values for the different methods


```{r, eval=TRUE, fig.width = 12}
p_value_hist(r )
```



## Compare the significant peptides



```{r, eval = TRUE}
compare_calls(r)
```

```{r,fig.width=6, fig.height=6}
comparisons <- data.frame(
  control = c('665e6428','665e6428'),
  c_seconds = c(0,150),
  treatment = c('665e6428','665e6428'),
  t_seconds = c(150, 0)
)
  
many <- compare_many(d, comparisons, metrics = c("bootstrap_t", "rank_product"))
many
```

```{r,eval=TRUE, fig.width=10}
plot_heatmap(many, metric = "bootstrap_t", log = TRUE,  col_order = c("665e6428_150-665e6428_0", "665e6428_0-665e6428_150"))

plot_heatmap(many, metric = "rank_product", log = TRUE)
```

```{r, fig.width=10}
plot_heatmap(many, metric = "bootstrap_t", log = TRUE,  col_order = c("665e6428_150-665e6428_0", "665e6428_0-665e6428_150"), only_sig_points = FALSE)
```
